---
title: "Starting with R"
date: "Updated `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    # Run xaringan::summon_remark() for this
    #chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE) 
knitr::opts_chunk$set(echo = FALSE, dev.args = list(png = list(type = "cairo")), warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(directlabels)
library(ggdag)
theme_metro <- function(x) {
  theme_minimal() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16),
        axis.title.x = element_text(hjust = 1),
        axis.title.y = element_text(hjust = 1, angle = 0))
}
theme_metro_regtitle <- function(x) {
  theme_minimal() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
```

# Welcome to R

- So cruel to set you up for econometrics and take a detour
- But we're going to spend some time today focusing on getting ourselves set up to use R
- R is a programming language designed for use in statistics
- We will be using it throughout the course

---

# Getting Set Up

- We can install R at [R-project.org](R-project.org)
- Then we can download and install RStudio at [RStudio.com](Rstudio.com)
- Then, R is installed and ready to go! Open up RStudio to get going
- Alternately, you can skip installing anything and use [RStudio.cloud](RStudio.cloud) which runs entirely in the browser but is a bit slower and requires a (free) account

---

# The RStudio Panes

- Console
- Environment Pane
- Browser Pane
- Source Editor
- Li'l tip: Ctrl/Cmd + Shift + number will maximize one of these panes

---

# Console

- Typically bottom-left
- This is where you can type in code and have it run immediately
- Or, when you run code from the Source Editor, it will show up here
- It will also show any output or errors

---

# Console Example

- Let's copy/paste some code in there to run

```{r, echo = TRUE, eval = FALSE}
#Generate 500 heads and tails
data <- sample(c("Heads","Tails"),500,replace=TRUE)
#Calculate the proportion of heads
mean(data=="Heads")
#This line should give an error - it didn't work!
data <- sample(c("Heads","Tails"),500,replace=BLUE)
#This line should give a warning
#It did SOMETHING but maybe not what you want
mean(data)
#This line won't give an error or a warning
#But it's not what we want!
mean(data=="heads")
```

---

# What We Get Back

- We can see the code that we've run
- We can see the output of that code, if any
- We can see any errors or warnings (in <span style="color: red">red</span>). Remember - errors mean it didn't work. Warnings mean it *maybe* didn't work.
- Just because there's no error or warning doesn't mean it DID work! Always think carefully

---

# Environment Tab

- Environment tab shows us all the objects we have in memory
- For example, we created the `data` object, so we can see that in Environment
- It shows us lots of handy information about that object too
- (we'll get to that later)
- You can erase everything with that little broom button (technically this does `rm(list=ls())`)

---

# Browser Pane

- Bottom-right
- Lots of handy stuff here!
- Mostly, the *outcome* of what you do will be seen here
- Plots you make will show up here
- Some functions create tables or output that show up in Viewer
- Packages tab - avoid for loading, but the update button is nice!

---

# Files Tab

- Basic file browser
- Handy for opening up files
- Can also help you set the working directory:
    - Go to folder
    - In menu bar, Session
    - Set Working Directory
    - To Files Pane Location

---

# Help Tab

- This is where help files appear when you ask for them
- You can use the search bar here, or `help()` in the console
- Of course, plenty of materials also available for whatever on Google!

---

# Source Pane

- You should be working with code FROM THIS PANE, not the console!
- Why? Replicability!
- Also, COMMENTS! USE THEM! PLEASE! `#` lets you write a comment.
- Switch between tabs like a browser

---

# Running Code from the Source Pane

- Select a chunk of code and hit the "Run" button
- Click on a line of code and do Ctrl/Cmd-Enter to run just that line and advance to the next <- Super handy!
- Going one line at a time lets you check for errors more easily
- Let's try some!

```{r, eval=FALSE, echo=TRUE}
data(mtcars)
mean(mtcars$mpg)
mean(mtcars$wt)
372+565
log(exp(1))
2^9
(1+1)^9
```

---

# Autocomplete

- RStudio comes with autocomplete! 
- Typing in the Source Pane or the Console, it will try to fill in things for you
    - Command names (shows the syntax of the function too!)
    - Object names from your environment
    - Variable names in your data
- Let's try redoing the code we just did, typing it out

---

# Help

- Autocomplete is one way that RStudio tries to help you out
- The way that R helps you out is with the documentation
- When you start doing anything serious with a computer, like programming, the most important skills are:
     - Knowing to read documentation
     - Knowing to search the internet for help (always!)
 
---
     
# help()

- You can get the documentation on most R objects using the `help()` function
- `help(mean)`, for example, will show you:
    - What the function is
    - The "syntax" for the function and the order the arguments go in
    - The available options for the function
    - Other, related functions, like `weighted.mean`
    - Ideally, some examples of proper use
- Not just for functions/commands - some data sets will work too! Try `help(mtcars)`

---

# Packages

- R runs on user-contributed packages that contain functions you can use
- Packages stored on CRAN can be installed with `install.packages('packagename')`
- Once a package is installed you don't need to install it again (except to update it)
- But every time you open R you'll need to load it in again with `library(packagename)` if you want to use its functions

```{r, eval = FALSE}
# If we haven't installed it yet
# install.packages('vtable')
library(vtable)
vtable(iris)
```

---

# RMarkdown

- Go File $\rightarrow$ New File $\rightarrow$ RMarkdown to create a new RMarkdown document
- A text document with some basic layout, for example hashtags for sectioning
- Include formatted code with backticks
- Backtick followed by `r` to make it *execute*
- Triple backticks for not-inline code chunk. Let's look!

---

# Working in R

- Everything in R is an object
- We can only really do three things in R:
     - Create objects with `<-` or `=`
     - Send objects through functions to manipulate them
     - Look at objects

---

# Objects

- Let's create a basic object

```{r, echo = TRUE}
a <- 1
```

- We've taken `1` and stored it inside the `a` object
- Now if we just type `a` by itself, it will show us the `1` we stored inside
- This is a numeric object. We could also have `'strings'` or logicals: `TRUE` or `FALSE` or factors

---

# Objects

- We can manipulate objects

```{r, echo = TRUE}
a + 1
```

- "create a new object that takes `a` (1) and adds 1 to it (`1+1=2`)
- Notice that `a` itself doesn't change until we *reassign it*

```{r, echo = TRUE}
a
a <- a + 1
a
```

---

# Vectors

- A vector is a collection of objects of the same type
- While lots of functions create vectors, we can also make them ourselves with `c()` (concatenate)

```{r, echo = TRUE}
my_vector <- c(1,8,2,4,3)
```

- We can refer to a certain element of a vector with square brackets `[]` with a single number or a range `start:end`

```{r, echo = TRUE}
my_vector[3:4]
```

- Or using another vector of logicals (`TRUE` and `FALSE`) to pick elements (handy when we get to data!)

```{r, echo = TRUE}
my_vector[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
```

---

# Data Frames

- A `data.frame` is what we'll be working with most of the time.
- It's a collection of vectors of the same length
- We can create them ourselves, e.g. `data.frame(a = c(1,5,3,2), b = c('a','d','b','f'))`, but often we will read in a file or use `data()` to get a data set

```{r, echo = TRUE}
data(mtcars)
mtcars
```

---

# Data Frames

- We can get a vector back out of the data frame with `$` or `[[]]`

```{r, echo = TRUE}
mtcars$cyl[1:5]
mtcars[['cyl']][1:5]
```

---

# Data Frames

- Which we might want to do to send it to a function like `mean` that takes a vector!

```{r, echo = TRUE}
mean(mtcars$cyl)
```

---

# dplyr

- In this course, we'll be working with data largely using the **dplyr** package, which is a part of the **tidyverse**
- **dplyr** is a collection of verbs for manipulating data, all strung together with the pipe `%>%` (which technically **dplyr** just borrows from **magrittr**)
- **dplyr** has lots of functions in it. Today we'll cover just five: `pull()`, `filter()`, `%>%`, `select()`, and `mutate()`.
- For all of these, don't forget that if you want to take the data frame you've changed and *keep* that change, you have to re-assign the object with `<-` or `=`!

---

# pull()

- Pull just takes a vector back out of a data set, just like `$` or `[[]]`
- So why use it? It plays well with the pipe (upcoming)
- Also, it's flexible. It takes a variable name, a variable name as a string, or a column number

```{r, echo = TRUE, eval = FALSE}
mtcars$cyl
pull(mtcars, cyl)
pull(mtcars, 'cyl')
pull(mtcars, 2)
```

---

# filter()

- `filter()` picks just some of the *rows* of your data
- Important for analyzing a subset of your data!
- Give it a logical statement that's `TRUE` for the rows you want. `==` checks for equality!

```{r, echo = TRUE}
filter(mtcars, cyl == 4 & am == 1)
```

---

# The pipe

- The pipe `%>%` is important for making your code readable, and minimizing balanced-parentheses errors
- It takes whatever is on its left and makes it the first argument of the function on the right
- So whatever object you're working with you take, ship it along to the next function, process, then ship along again, then ship along again! Like a conveyer belt
- Notice that all **dplyr** functions take the data frame as the first argument, making it easy to chain them
- "Ships along" anything, including vectors or single numbers, not just data frames! Track what the object being shipped is in each step.


---

# The pipe

- See how clean it can make the code!

```{r, echo = TRUE}
mean(mtcars[mtcars$am == 1,]$cyl, na.rm = TRUE)
```

vs.

```{r, echo = TRUE}
mtcars %>%
  filter(am == 1) %>%
  pull(cyl) %>%
  mean(na.rm = TRUE)
```

---

# select()

- `select()` is like `filter()`, but instead of picking rows it picks columns
- Unlike `pull()` it doesn't give you a vector - it gives you back a data frame with fewer columns

```{r, echo = TRUE}
mtcars %>%
  select(mpg, cyl) %>%
  summary()
```

---

# mutate()

- `mutate()` creates a new column using the old ones. You can assign multiple columns at once!
- The syntax is `NewVariableName = FunctionOfOldVariables`
- Don't forget to save the object!

```{r, echo = TRUE}
mtcars <- mtcars %>%
  mutate(high_mpg = mpg > median(mpg))
mtcars %>%
  pull(high_mpg) %>%
  table()
```

---

# Basics of R

- This has been a lot of information!
- This is mostly to show you what's there. To really learn it you'll have to get used to applying it yourself
- For that we will be using the **swirl** package which will walk us through using these commands!

---

# Swirl

- Open up R and install Swirl with `install.packages('swirl')`
- Then, load up swirl with `library(swirl)`
- Download the Swirl course for this class (the first time you use it) with `install_course_github('NickCH-K','Econometrics')`
- Then start swirl with `swirl()` and pick Econometrics, and the R Basics lesson. Let's work through this!

