---
title: "Within Variation and Fixed Effects"
subtitle: "i.e. one thing to do when measurement eludes you"
date: "Updated `r Sys.Date()`"
output:
  xaringan::moon_reader:
    self_contained: TRUE
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    # Run xaringan::summon_remark() for this
    #chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE) 
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(gganimate)
library(fixest)
library(magick)
library(wooldridge)
library(dagitty)
library(ggdag)
library(scales)
library(Cairo)
theme_metro <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16),
        axis.title.x = element_text(hjust = 1),
        axis.title.y = element_text(hjust = 1, angle = 0))
}
theme_void_metro <- function(x) {
  theme_void() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
theme_metro_regtitle <- function(x) {
  theme_classic() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
```

# Check-in

- So far we've been learning about how to set up, run, and interpret an ordinary least squares regression
- This is a key skill for anyone doing anything with data - even if you never run a regular ol' linear regression again, pretty much everything else in applied stats builds off of it in some way
- Another thing we've been doing is thinking about how to design and add controls to that regression to *identify* our effect of interest by closing back doors

---

# The Measurement Problem...

- And this has led us to some issues that have already popped up!
- For this approach to work, we need to not only *figure out* what we need to control for, using our diagram, but we need to *actually control for it*
- A lot of the time we don't have that data!
- And thus all the skeptical comments we had about the designs we came up with

---

# A Pickle

- So obviously this is a problem, and it's not one we can reason or trick our way out of
- If we don't have the variable we need to control for, we don't have it
- ... or do we?

---

# The Rest of the Term

- Much of the rest of the term is going to be focused on *finding ways to control for stuff that we can't measure*
- Seems impossible! But it is possible, at least in some circumstances
- Today, we will be talking about *within variation* and *between variation*, and the ability to control for all *between variation* using *fixed effects*

---

# Panel Data

- We are working now in the domain of *panel data*
- Panel data is when you observe the same individual over multiple time periods
- "Individual" could be a person, or a company, or a state, or a country, etc. There are $N$ individuals in the panel data
- "Time period" could be a year, a month, a day, etc.. There are $T$ time periods in the data
- For now we'll assume we observe each individual the same number of times, i.e. a *balanced* panel (so we have $N\times T$ observations)
- You can use this stuff with unbalanced panels too, it just gets a little more complex

---

# Panel Data

- Here's what (a few rows from) a panel data set looks like - a variable for individual (county), a variable for time (year), and then the data

```{r, dev = 'CairoPNG'}
data(crime4)
crime4 %>%
  select(county, year, crmrte, prbarr) %>%
  rename(County = county,
         Year = year,
         CrimeRate = crmrte,
         ProbofArrest = prbarr) %>%
  slice(1:9) %>%
  knitr::kable(note = '...') %>%
  kableExtra::add_footnote('9 rows out of 630. "Prob. of Arrest" is estimated probability of being arrested when you commit a crime', notation = 'none')
```

---

# Between and Within

- Let's pick a few counties and graph this out

```{r, dev = 'CairoPNG'}
crime4 %>% 
  filter(county %in% c(1,3,7, 23),
         prbarr < .5) %>%
  group_by(county) %>%
  mutate(label = case_when(
    crmrte == max(crmrte) ~ paste('County',county),
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x =  prbarr, y = crmrte, color = factor(county), label = label)) + 
  geom_point() + 
  geom_text(hjust = -.1, size = 14/.pt) + 
  theme_metro_regtitle() + 
  labs(x = 'Probability of Arrest', 
       y = 'Crime Rate',
       caption = 'One outlier eliminated in County 7.') + 
  #scale_x_continuous(limits = c(.15, 2.5)) + 
  guides(color = FALSE, label = FALSE) + 
  scale_color_manual(values = c('black','blue','red','purple'))
```

---

# Between and Within

- If we look at the overall variation, just pretending this is all together, we get this

```{r, dev = 'CairoPNG'}
crime4 %>% 
  filter(county %in% c(1,3,7, 23),
         prbarr < .5) %>%
  group_by(county) %>%
  mutate(label = case_when(
    crmrte == max(crmrte) ~ paste('County',county),
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x =  prbarr, y = crmrte, color = factor(county), label = label)) + 
  geom_point() + 
  geom_text(hjust = -.1, size = 14/.pt) + 
  theme_metro_regtitle() + 
  labs(x = 'Probability of Arrest', 
       y = 'Crime Rate',
       caption = 'One outlier eliminated in County 7.') + 
  #scale_x_continuous(limits = c(.15, 2.5)) + 
  guides(color = FALSE, label = FALSE) + 
  scale_color_manual(values = c('black','blue','red','purple')) + 
  geom_smooth(method = 'lm', aes(color = NULL, label = NULL), se = FALSE)
```


---


# Between and Within

- BETWEEN variation is what we get if we look at the relationship between the *means of each county*

```{r, dev = 'CairoPNG'}
crime4 %>% 
  filter(county %in% c(1,3,7, 23),
         prbarr < .5) %>%
  group_by(county) %>%
  mutate(label = case_when(
    crmrte == max(crmrte) ~ paste('County',county),
    TRUE ~ NA_character_
  ),
  mcrm = mean(crmrte),
  mpr = mean(prbarr)) %>%
  ggplot(aes(x =  prbarr, y = crmrte, color = factor(county), label = label)) + 
  geom_point() + 
  geom_text(hjust = -.1, size = 14/.pt) + 
  theme_metro_regtitle() + 
  labs(x = 'Probability of Arrest', 
       y = 'Crime Rate',
       caption = 'One outlier eliminated in County 7.') + 
  #scale_x_continuous(limits = c(.15, 2.5)) + 
  guides(color = FALSE, label = FALSE) + 
  scale_color_manual(values = c('black','blue','red','purple')) + 
  geom_point(aes(x = mpr, y = mcrm), size = 20, shape = 3, color = 'darkorange') + 
  annotate(geom = 'text', x = .3, y = .02, label = 'Means Within Each County', color = 'darkorange', size = 14/.pt)
```

---

# Between and Within

- And I mean it! Only look at those means! The individual year-to-year variation within county doesn't matter.

```{r, dev = 'CairoPNG'}
crime4 %>% 
  filter(county %in% c(1,3,7, 23),
         prbarr < .5) %>%
  group_by(county) %>%
  mutate(label = case_when(
    crmrte == max(crmrte) ~ paste('County',county),
    TRUE ~ NA_character_
  ),
  mcrm = mean(crmrte),
  mpr = mean(prbarr)) %>%
  ggplot(aes(x =  prbarr, y = crmrte, color = factor(county), label = label)) + 
  #geom_point() + 
  #geom_text(hjust = -.1, size = 14/.pt) + 
  theme_metro_regtitle() + 
  labs(x = 'Probability of Arrest', 
       y = 'Crime Rate',
       caption = 'One outlier eliminated in County 7.') + 
  #scale_x_continuous(limits = c(.15, 2.5)) + 
  guides(color = FALSE, label = FALSE) + 
  scale_color_manual(values = c('black','blue','red','purple')) + 
  geom_point(aes(x = mpr, y = mcrm), size = 20, shape = 3, color = 'darkorange') + 
  geom_smooth(aes(color = NULL), method = 'lm', se = FALSE)+
  annotate(geom = 'text', x = .3, y = .02, label = 'OLS Fit on These Four Points', color = 'blue', size = 14/.pt)
```

---

# Between and Within

- Within variation goes the other way - it treats those orange crosses as their own individualized sets of axes and looks at variation *within* county from year-to-year only!
- We basically slide the crosses over on top of each other and then analyze *that* data

```{r, echo=FALSE, fig.width=5, fig.height=4.5}
cranim <- crime4 %>% 
  filter(county %in% c(1,3,7, 23),
         prbarr < .5) %>%
  mutate(allcrm = mean(crmrte),
         allmpr = mean(prbarr)) %>%
  group_by(county) %>%
  mutate(label = case_when(
    crmrte == max(crmrte) ~ paste('County',county),
    TRUE ~ NA_character_
  ),
  mcrm = mean(crmrte),
  mpr = mean(prbarr),
  stage = '1. Raw Data')
cranim <- cranim %>%
  bind_rows(cranim %>% 
              mutate(crmrte = crmrte - mcrm + allcrm,
                     prbarr = prbarr - mpr + allmpr,
                     mcrm = allcrm,
                     mpr = allmpr,
                     stage = '2. Remove all between variation'))

p <- ggplot(cranim, aes(x =  prbarr, y = crmrte, color = factor(county), label = label)) + 
  geom_point() + 
  geom_text(hjust = -.1, size = 14/.pt)  + 
  labs(x = 'Probability of Arrest', 
       y = 'Crime Rate',
       caption = 'One outlier eliminated in County 7.') + 
  #scale_x_continuous(limits = c(.15, 2.5)) + 
  guides(color = FALSE, label = FALSE) + 
  scale_color_manual(values = c('black','blue','red','purple')) + 
  geom_smooth(aes(color = NULL), method = 'lm', se = FALSE)+
  geom_point(aes(x = mpr, y = mcrm), size = 20, shape = 3, color = 'darkorange') + 
  transition_states(stage) + 
  theme_metro_regtitle()

animate(p, nframes = 80)

```

---

# Between and Within

- We can clearly see that *between counties* there's a strong positive relationship
- But if you look *within* a given county, the relationship isn't that strong, and actually seems to be negative
- Which would make sense - if you think your chances of getting arrested are high, that should be a deterrent to crime
- But what are we actually doing here? Let's think about the causal diagram / data-generating process!
- What goes into the probability of arrest and the crime rate? Lots of stuff!

---

# The Crime Rate

- "LocalStuff" is just all the things unique to that area
- "LawAndOrder" is how committed local politicians are to "Law and Order Politics"

```{r, dev = 'CairoPNG'}
dag <- dagify(CrimeRate ~ ProbArrest + Poverty + CivilRights + LocalStuff,
              ProbArrest ~ Police + LawAndOrder + CivilRights + Poverty + LocalStuff,
              Police ~ LawAndOrder,
              CivilRights ~ LawAndOrder,
              Poverty ~ LocalStuff,
              coords=list(
                x=c(ProbArrest = 1, CrimeRate = 4, Poverty = 2.5, Police = 1.5, LawAndOrder = 1, CivilRights = 2.5, LocalStuff = 4),
                y=c(ProbArrest = 1, CrimeRate = 1, Poverty = 2, Police = 1.5, LawAndOrder = 2, CivilRights = 1.25, LocalStuff = 2)
              )) %>% tidy_dagitty()
ggdag_classic(dag,node_size=10) + 
  theme_void_metro() + 
  expand_limits(x=c(.5,4.5))

```


---

# Between and Within

- For each of these variables we can ask if they vary *between groups* and/or *within groups*
- LocalStuff is all the stuff unique to that county - geography, landmarks, the quality of the schools, almost by definition this only varies *between groups*. It's not like the things that make your county unique are different each year (or at least not very different)
- Whether the county has LawAndOrder and how many CivilRights you have might change a bit year to year, but in general, political climates like that change pretty slowly. At a bit of a stretch we can call that something that only varies between groups too
- Police budgets (and thus number of police on the streets) and Poverty (which varies with the economy) vary both between counties, but also *within* counties from year to year
- Variables with between variation only (by our assumption): LocalStuff, LawAndOrder, CivilRights
- Variables with both between and within variation: Police, Poverty

---

# Between and Within

- Let's simplify our graph!
- Some of the variables only vary *between counties*
- So, we can replace those variables on the graph with the variable County
- Right? That's where all the variation is anyway

---

# The Crime Rate

- "LocalStuff" is just all the things unique to that area
- "LawAndOrder" is how committed local politicians are to "Law and Order Politics"

```{r, dev = 'CairoPNG'}
dag <- dagify(CrimeRate ~ ProbArrest + Poverty + County,
              ProbArrest ~ Police + County + Poverty,
              Police ~ County,
              Poverty ~ County,
              coords=list(
                x=c(ProbArrest = 1, CrimeRate = 4, Poverty = 1, Police = 1.5, County = 3),
                y=c(ProbArrest = 1, CrimeRate = 1, Poverty = 2, Police = 1.5, County = 2)
              )) %>% tidy_dagitty()
ggdag_classic(dag,node_size=10) + 
  theme_void_metro() + 
  expand_limits(x=c(.5,4.5))

```

---

# Between and Within

- Now the task of identifying ProbArrest $\rightarrow$ CrimeRate becomes much simpler!
- If we control for County, that will close a lot of back doors for us
- (based on the diagram, all we need to control for is County and Poverty!)
- Conveniently, we can control for County just like it was any other variable!
- And when we do, we automatically *control for all variables that only have between variation*, whatever they are, even if we can't measure them directly or didn't think about them
- *All that's left is the within variation*


---

# Concept Checks

- For each of these variables, would we expect them to have within variation, between variation, or both?
- (Individual = person) How a child's height changes as they age.
- (Individual = person) In a data set tracking many people over many years, the variation in the number of children a person has over their lifetime.
- (Individual = city) In any given year, Paris, France will have more restaurants than Paris, Texas.

---

# Removing Between Variation

- Okay so that's the concept
- Remove all the between variation so that all that's left is within variation
- And in the process control for any variables that are made up only of between variation
- How can we actually do this? And what's really going on?
- Let's first talk about the regression model itself that this implies
- Then let's actually do the thing. There are two main ways: *de-meaning* and *binary variables* (they give the same result, for balanced panels anyway)

---

# Estimation vs. Design

- To be clear, this is *exactly 0% different* from what we've done before in terms of controlling for stuff
- And in fact we're about to do the *exact same thing we did before* by just adding a categorical control variable for `county` or whatever
- (and in fact the "within" thing holds with other categorical controls - a categorical control for education isolates variation "within education levels")
- The difference is the *reason we're doing it*. It's fixed effects because *a categorical control for individual controls for a lot of stuff*, and we think closes a *lot* of back doors for us, not just one, and not just ones we can measure


---


# The Model


The $it$ subscript says this variable varies over individual $i$ and time $t$

$$Y_{it} = \beta_0 + \beta_1 X_{it} + \varepsilon_{it}$$

- What if there are individual-level components in the error term causing omitted variable bias? 
- $X_{it}$ is related to LocalStuff which is not in the model and thus in the error term!
- Regular ol' omitted variable bias. If we don't adjust for the individual effect, we get a biased $\hat{\beta}_1$ 
- (this bias is called "pooling bias" although it's really just a form of omitted variable bias)
- We really have this then:

$$Y_{it} = \beta_0 + \beta_1 X_{it} + (\alpha_i + \varepsilon_{it})$$

---

# De-meaning

- Let's do de-meaning first, since it's most closely and obviously related to the "removing between variation" explanation we've been going for
- The process here is simple!

1. For each variable $X_{it}$, $Y_{it}$, etc., get the mean value of that variable for each individual $\bar{X}_i, \bar{Y}_i$
2. Subtract out that mean to get residuals $(X_{it} - \bar{X}_i), (Y_{it} - \bar{Y}_i)$
3. Work with those residuals

- That's it!

---

# How does this work?

- That $\alpha_i$ term gets absorbed
- The residuals are, by construction, no longer related to the $\alpha_i$, so it no longer goes in the residuals!

$$(Y_{it} - \bar{Y}_i) = \beta_0 + \beta_1(X_{it} - \bar{X}_i) + \varepsilon_{it}$$

---

# Let's do it!

- We can use `group_by` to get means-within-groups and subtract them out

```{r, echo = TRUE}
data(crime4, package = 'wooldridge')
crime4 <- crime4 %>%
  # Filter to the data points from our graph
  filter(county %in% c(1,3,7, 23),
         prbarr < .5) %>%
  group_by(county) %>%
  mutate(mean_crime = mean(crmrte),
         mean_prob = mean(prbarr)) %>%
  mutate(demeaned_crime = crmrte - mean_crime,
         demeaned_prob = prbarr - mean_prob)
```

---

# And Regress!

```{r, echo = TRUE}
orig_data <- feols(crmrte ~ prbarr, data = crime4)
de_mean <- feols(demeaned_crime ~ demeaned_prob, data = crime4)
etable(orig_data, de_mean)
```

---

# Interpreting a Within Relationship

- How can we interpret that slope of `-0.03`?
- This is all *within variation* so our interpretation must be *within-county*
- So, "comparing a county in year A where its arrest probability is 1 (100 percentage points) higher than it is in year B, we expect the number of crimes per person to drop by .03"
- Or if we think we've causally identified it (and want to work on a more realistic scale), "raising the arrest probability by 1 percentage point in a county reduces the number of crimes per person in that county by .0003".
- We're basically "controlling for county" (and will do that explicitly in a moment)
- So your interpretation should think of it in that way - *holding county constant* i.e. *comparing two observations with the same value of county* i.e. *comparing a county to itself at a different point in time*

---

# Concept Checks

- Why does subtracting the within-individual mean of each variable "control for individual"?
- In a sentence, interpret the slope coefficient in the estimated model $(Y_{it} - \bar{Y}_i) = 2 + 3(X_{it} - \bar{X}_i)$ where $Y$ is "blood pressure", $X$ is "stress at work", and $i$ is an individual person

---

# The Least Squares Dummy Variable Approach

- De-meaning the data isn't the only way to do it!
- You can also use the least squares dummy variable (another word for "binary variable") method
- We just treat "individual" like the categorical variable it is and add it as a control! Again, the regression approach is exactly the same as with any categorical control, but the *research design* reason for doing it is different

---

# Let's do it!

```{r, echo = TRUE}
lsdv <- feols(crmrte ~ prbarr + factor(county), data = crime4)
etable(orig_data, de_mean, lsdv, keep = c('prbarr', 'demeaned_prob'))
```

---

# The same!

- The result is the same, as it should be
- Except for that $R^2$ - What is that "within R2"?
- Because de-meaning takes out the part explained by the fixed effects ( $\alpha_i$ ) *before* running the regression, while LSDV does it *in* the regression
- So the .94 is the portion of `crmrte` explained by `prbarr` *and* `county`, whereas the .21 is the "within - $R^2$ " - the portion of *the within variation* that's explained by `prbarr`
- Neither is wrong (and the .94 isn't "better"), they're just measuring different things

---

# Why LSDV?

- A benefit of the LSDV approach is that it calculates the fixed effects $\alpha_i$ for you
- We left those out of the table with the `coefs` argument of `export_summs` (we rarely want them) but here they are:

```{r}
lsdv
```

- Interpretation is exactly the same as with a categorical variable - we have an omitted county, and these show the difference relative to that omitted county

---

# Why LSDV?

- This also makes clear another element of what's happening! Just like with a categorical var, the line is moving *up and down* to meet the counties
- Graphically, de-meaning moves all the points together in the middle to draw a line, while LSDV moves the line up and down to meet the points

```{r, dev = 'CairoPNG'}
crime4 %>%
  ungroup() %>%
  mutate(pred = predict(lsdv)) %>%
  group_by(county) %>%
  mutate(label = case_when(
    crmrte == max(crmrte) ~ paste('County',county),
    TRUE ~ NA_character_
  )) %>%
  ggplot(aes(x =  prbarr, y = crmrte, color = factor(county), label = label)) + 
  geom_point() + 
  geom_text(hjust = -.1, size = 14/.pt) + 
  geom_line(aes(y = pred, group = county), color = 'blue') +
  theme_metro_regtitle() + 
  labs(x = 'Probability of Arrest', 
       y = 'Crime Rate',
       caption = 'One outlier eliminated in County 7.') + 
  #scale_x_continuous(limits = c(.15, 2.5)) + 
  guides(color = FALSE, label = FALSE) + 
  scale_color_manual(values = c('black','blue','red','purple'))
```

---

# Why Not LSDV?

- LSDV is computationally expensive
- If there are a lot of individuals, or big data, or if you have many sets of fixed effects (yes you can do more than just individual - we'll get to that next time!), it can be very slow
- Most professionally made fixed-effects commands use de-meaning, but then adjust the standard errors properly
- (They also leave the fixed effects coefficients off the regression table by default)

---

# Going Professional

- Applied researchers rarely do either of these, and rather will use a command specifically designed for fixed effects
- Like good ol' `feols()`! (what did you think the "fe" part stood for?)
- Note there are also functions in **fixest** that do fixed effects in non-linear models like logit, probit, or poisson regression (`feglm()` and `fepois()`)
- Plus, it clusters the standard errors by the first fixed effect by default, which we usually want!

---

# Going Professional

```{r, echo = TRUE}
library(fixest)
pro <- feols(crmrte ~ prbarr | county, data = crime4)
etable(de_mean, pro)
```

---

# Limits to Fixed Effects

- Okay! At this point we have the concept behind fixed effects, can execute them, and know what they're good for
- What aren't they good for?

1. They don't control for anything that has within variation
2. They control away *everything* that's between-only, so we can't see the effect of anything that's between-only ("effect of geography on crime rate?" Nope!)
3. Anything with only a *little* within variation will have most of its variation washed out too ("effect of population density on crime rate?" probably not)
4. The estimate pays the most attention to individuals with *lots of variation in treatment*

- 2 and 3 can be addressed by using "random effects" instead but we aren't covering that in this class (see the The Effect chapter on Fixed Effects for more)

---

# Concept Checks

- Why can't we use individual-person fixed effects to study the impact of race on traffic stops?
- The within $R^2$ from is .3, and the overall $R^2$ is .5. Interpret these two numbers in sentences
- In a sentence, interpret the slope coefficient in the estimated model $(Y_{it} - \bar{Y}_i) = 1 + .5(X_{it} - \bar{X}_i)$ where $Y$ is "school funding per child" and $X$ is "population growth", and $i$ is city

---

# Swirl

- Open up the Fixed Effects Swirl and let's do it!

