---
title: "Applying Hypothesis Testing"
subtitle: "Part 2: How to Do It"
date: "Updated `r Sys.Date()`"
output:
  xaringan::moon_reader:
    self_contained: TRUE
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    # Run xaringan::summon_remark() for this
    #chakra: libs/remark-latest.min.js
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE) 
knitr::opts_chunk$set(echo = FALSE, dev.args = list(png = list(type = "cairo")), warning = FALSE, message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(jtools)
library(scales)
library(car)
# remotes::install_github('hadley/emo')
library(emo)
theme_metro <- function(x) {
  theme_minimal() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16),
        axis.title.x = element_text(hjust = 1),
        axis.title.y = element_text(hjust = 1, angle = 0))
}
theme_metro_regtitle <- function(x) {
  theme_minimal() + 
  theme(panel.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        plot.background = element_rect(color = '#FAFAFA',fill='#FAFAFA'),
        text = element_text(size = 16))
}
```

# Recap

- We are trying to characterize the *uncertainty* in our estimate that comes from sampling variation
- One way to do this is to characterize the sampling distribution: show the standard error and construct a confidence interval
- Another is if we have a *null hypothesis* of interest - for parameters giving relationships like slopes, usually that the parameter is 0 (no relationship)
- Then we can look at the sampling distribution *assuming the null is true* and see if our actual result is too weird to believe; if it is, reject the null!

---

# The Null Distribution

- Hypothesis testing centers around the concept of the sampling distribution and, further, the null distribution
- We've talked so far about parameters that have a normal sampling distribution, i.e. follow a normal distribution
- This is a key assumption to have made
- We have to have an idea of what the null distribution of our parameter *is* in order to figure out *how weird* our result is

---

# The Normal Null

- Let's talk about the normal distribution some more to get more intuition on what's happening, then branch out. 
- When we have an estimate, like $\hat{\beta}_1$, that follows a normal distribution, we know that the sampling distribution:
    - Is symmetric
    - Can be wider or narrower depending on the standard error
    - Can be transformed into a "standard normal" (mean 0 and standard deviation 1) by subtracting the mean and dividing by the standard deviation

---

# The Normal Null

- Because of that last point, when evaluating an estimate like $\hat{\beta}_1$ we will transform it into a *Z-score*, which is

$$\frac{\hat{\beta}_1 - Null}{s.e.(\hat{\beta_1})}$$

- That is, we subtract our null hypothesis value and divide by the standard error
- This Z-score is our *test statistic*

---

# The Normal Null

- Why do this transformation? Because the *sampling distribution assuming the null is true* is a normal with mean $Null$ and standard error $s.e.(\beta_1)$
- So by subtracting the null (mean of the sampling distribution) and dividing by $s.e.(\beta_1)$ (standard deviation of the sampling distribution), the distribution under the null becomes a normal with mean-0 and standard-deviation-1, the standard normal. Very easy to work with!
- The sampling distribution *of $\hat{\beta}_1$* under the null was the original normal distribution we started with, but the sampling distribution *of the Z-Score* under the null is a normal with mean 0 and s.d. 1
- So if 2.5% of the area of the distribution is above our original estimate of $4.92$ in the original null distribution with mean $1$ and s.e. $2$, then 2.5% of the area will also be above our Z-score of $(4.92-1)/2$ in the standard normal


---

# The Normal Null

- This brings us to the concept of *critical values*
- When we pick an $\alpha$, we will reject the null if the p-value is below $\alpha$
- This happens if $\hat{\beta}_1$ is a certain distance away from the null or farther
- If we use the *same distribution every time* (standard null), that distance will always be the same!
- So instead of checking the percentage under the distribution every time, we just figure out a critical value and see if we're farther away from that. Much easier!

---

# The Normal Null

- For the standard normal, 5% is outside the bounds of $Z = -1.96$ and $Z = 1.96$ (2.5% on each side - a two-sided test!) Remember this number?
- For 10% it's 1.65 and for 1% it's 2.58. How much area is in each shaded part?

```{r}
df <- data.frame(x = -300:300/100, y = dnorm(-300:300/100))

ggplot(df, aes(x = x, y = y)) + geom_line() + 
  scale_x_continuous(breaks = qnorm(c(.05,.025,.005,.95,.975,.995)), labels = number) + 
  labs(x = "X", y = "Density") +
  theme_metro() + 
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = -20)) + 
  geom_ribbon(data = filter(df, x <= qnorm(.05) | x > qnorm(.95)), mapping = aes(x = x, ymin = 0, ymax = y, group = x > 0),
              fill = 'red', alpha = .5) +
  geom_ribbon(data = filter(df, x <= qnorm(.025) | x > qnorm(.975)), mapping = aes(x = x, ymin = 0, ymax = y, group = x > 0),
              fill = 'blue', alpha = .5) + 
  geom_ribbon(data = filter(df, x <= qnorm(.005) | x > qnorm(.995)), mapping = aes(x = x, ymin = 0, ymax = y, group = x > 0),
              fill = 'orange')
```

---

# The Normal Null

- This critical-value approach means that we can check results much more easily
- Look at this regression table regressing number of house listings on "months it would take to sell this inventory". We can eyeball a test with a null of 0 at the 95% confidence level by calculating the test statistic in our heads and comparing to 1.96 (or 2 to be even easier) - when the null is 0, the Z-score is just $\hat{\beta}_1/s.e.(\hat{\beta})$

```{r}
data(txhousing)
housingmodel <- lm(listings ~ inventory, data = txhousing)
export_summs(housingmodel)
```

---

# Concept Checks

- Why do we transform $\hat{\beta}_1$ into a Z-score?
- When we transform $\hat{\beta}_1$, why do we have to use the null hypothesis value in our calculation?
- Why do we have to use the standard error?
- Is the coefficient on X in the below table significant at the 95% confidence level? Eyeball it!

```{r}
tib2 <- tibble(x = runif(20)) %>% mutate(y = runif(20) + x)
export_summs(lm(y~x, data = tib2), stars = NULL)
```

---

# The Normal Null

- So those are features of the normal null
- When do we use the normal null?
- Any time we have an *average* over *lots of observations*, the normal distribution pops up
- (it's not obvious, but OLS coefficients are averages)
- So for OLS coefficients, we just need "a lot of observations" to use a normal

---

# When Not to Use the Normal

- Small samples (how small is "small?" 0-30 is definitely small, 100+ probably not small, 30-100 gray area)
- Things that aren't means, like ratios
- In these cases, what can we do instead?

---

# The t-distribution

- The t-distribution is very similar to a normal distribution, except that it applies to means of smaller sample sizes, and instead of a mean and s.d. it has a number of degrees of freedom that determines how wide it is
- When you have fewer observations, you're more likely to get a mean that's far from the true mean, i.e. "fatter tails"

```{r}
dnt <- data.frame(x = -300:300/100, y = dnorm(-300:300/100), yt = dt(-300:300/100, 5), yt2 = dt(-300:300/100, 28))

ggplot(dnt, aes(x = x, y = y)) + 
  geom_line(color = 'red') + 
  geom_line(aes(x = x, y = yt), color = 'blue') + 
  geom_line(aes(x = x, y = yt2), color = 'orange') + 
  theme_metro() + 
  theme(panel.grid = element_blank()) +
  labs(x = "Value", y = "Density") + 
  annotate(geom = 'text', x = 1.5, y = .4, color = 'red', label = 'Standard normal', hjust = 0, size = 16/.pt) + 
  annotate(geom = 'text', x = 1.5, y = .375, color = 'blue', label = 't (dof = 5)', hjust = 0, size = 16/.pt) + 
  annotate(geom = 'text', x = 1.5, y = .35, color = 'orange', label = 't (dof = 28)', hjust = 0, size = 16/.pt)
```

---

# The t-distribution

- So, for smaller samples, it's a good idea to use the critical value from a t distribution than from a normal
- We do the same subtract-the-mean-and-divide-by-s.e. steps, just now we call it a t-score rather than a z-score
- How can we find the critical values of the t distribution? Don't bother memorizing, as they change depending on your sample size. Instead use a function that measures the proportion of the distribution weirder than a value, and take the inverse to get the critical value!

```{r, echo = TRUE}
qt(.025, df = 28)
```

- (psst... we could have also done that with the normal: `qnorm(.025) = ` `r qnorm(.025)`)

---

# Small Samples

- Using a null t-distribution when dealing with a small sample is important
- But it's not the only issue with small samples (even small samples large enough to use a normal!)
- The practice of hypothesis testing in general makes small samples more perilous!

---

# Small Samples

- Small samples reduce *power* considerably - it becomes difficult to reject a false null
- And also, the estimate in general will be much noisier!
- As samples get small, the number of *true rejections* drops as power drops, but the number of *super extreme noisy results* goes up!
- Think back to that animation where we had $N = 2$...
- So the smaller your sample, the better the chance that a given rejection of a null is a false positive rather than a true positive

---

# Small Samples

- Small samples also make it harder to detect small effects
- Power, false positive rates, and false negative rates rely not just on sample size and $var(X)$, but also *how big the true effect is that you are trying to find*
- Big effects are easy to find - $N = 2$ could tell you whether a parachute saves you from death
- Tiny effects need big samples to have power and estimate effects precisely - to have a 90% chance to reject the null of 0 effect for a pill that truly increases your IQ by .003 points you'd need a sample of about 8 billion

---

# Small Samples

- So in general:
    - Don't pay *too* much attention to studies with small samples
    - If a result is truly wild and unexpected, check if the sample is small - good chance it's just noise
    - If you do have to work with a small sample, maybe avoid hypothesis tests
    - But if you do use a test, use a t distribution null.

---

# F distribution

- Another null distribution that comes up a lot is the *F distribution*
- The F distribution is a distribution of *the ratio of two squared normal variables* (or the ratio of two sums of squared normal variables), a.k.a. the ratio of two $\chi^2$ distributed variables

```{r}
dfd <- data.frame(x = 0:300/100, y = df(0:300/100,1, 1), y2 = df(0:300/100, 1, 50), y3 = df(0:300/100, 5, 35))

ggplot(dfd, aes(x = x, y = y)) + 
  geom_line(color = 'red') + 
  geom_line(aes(x = x, y = y2), color = 'blue') + 
  geom_line(aes(x = x, y = y3), color = 'orange') + 
  theme_metro() + 
  theme(panel.grid = element_blank()) +
  labs(x = "Value", y = "Density") + 
  annotate(geom = 'text', x = 1.5, y = 1.1, color = 'red', label = 'F (df1 = 1, df2 = 1)', hjust = 0, size = 16/.pt) + 
  annotate(geom = 'text', x = 1.5, y = .9, color = 'blue', label = 'F (df1 = 1, df2 = 50)', hjust = 0, size = 16/.pt) + 
  annotate(geom = 'text', x = 1.5, y = .7, color = 'orange', label = 'F (df1 = 5, df2 = 35)', hjust = 0, size = 16/.pt)
```
---

# F distribution

- Why does this come up? Because it's useful for *comparing models*
- If we use a squared normal variable to measure some quality of a model (hint: OLS tries to minimize the sum of **squared** residuals; we can turn that into a measure of model fit, we'll get to that later)
- Then we can compare models by dividing one measure by the other
- If we surpass the critical value, then we can reject that the two models are equally good!

---

# F distribution

- Instead of a mean and a standard deviation, the F distribution is defined by two *degrees of freedom*, for the number of squared normals in the numerator and denominator, respectively
- When we're doing a comparison of models, these degrees of freedom will be based on *how many parameters* are being compared and the sample size
- We'll do more of that when we get to multivariate regression
- For now, we have one degree of freedom up top, and $N-1$ degrees of freedom on the bottom

---

# F distribution

- When we use the F distribution to do a test of a single regression coefficient, we're comparing the model with the variable included against the model without it included (i.e. $Y = \beta_0 + \beta_1X$ vs. $Y = \beta_0$ )
- The p-value will be exactly the same as if we'd used a normal null for a null hypothesis value of 0
- But, as mentioned, F distributions will be more handy when we get to multivariate regression

---

# F distribution

```{r}
hm2 <- lm(inventory~median,data=slice(txhousing,1:503))
export_summs(hm2, error_format = 'p-value = {p.value}')
paste('F statistic for the coefficient on median price is',number(linearHypothesis(hm2, 'median = 0')$F[2]), 'p-value =',number(linearHypothesis(hm2, 'median = 0')$`Pr(>F)`[2]))
```

---

# F distribution

- Testing one parameter with $496$ observations, so numerator df is $1$ and denominator is $496 - 1 = 495$

```{r}
dfd2 <- data.frame(x = 0:300/100, y = df(0:300/100,1, 495))

ggplot(dfd2, aes(x = x, y = y)) + 
  geom_line() + 
  geom_ribbon(data = filter(dfd2, x >= linearHypothesis(hm2, 'median = 0')$F[2]), aes(x = x, ymin = 0, ymax = y),
              alpha = .5) + 
  scale_x_continuous(breaks = linearHypothesis(hm2, 'median = 0')$F[2], labels = function(x) number(x, accuracy = .001)) +
  theme_metro() + 
  theme(panel.grid = element_blank()) +
  labs(x = "Value", y = "Density") + 
  annotate(geom = 'text', x = .82, y = .17, label = percent(linearHypothesis(hm2, 'median = 0')$`Pr(>F)`[2]), 
           hjust = 0, size = 16/.pt)
```

---

# Concept Checks

- Why do small samples increase the false negative rate (low power) but not increase the false positive rate?
- You want to test a regression coefficient estimated from a N = 2000 sample with a null hypothesis value of 1/2. What null distribution should you use?
- You want to test whether removing some variables from a regression model makes it worse. What null distibution should you use?
- You want to test if the mean of $X$ is equal to $3$ or not, from a sample of 25 observations. What test should you use, and how should you calculate the test statistic?

---

# Hypothesis Testing in R

- By default, regression results will perform tests using the t distribution
- Rather than switching to a normal null for big samples, it will keep using t, but with large samples that's basically normal anyway
- `summary(lm())` will show us the t-statistic as well as the p-value compared to 0
- `export_summs()` (which will properly adjust the tests for heteroskedasticity if you ask!) will show only standard errors by default, but we can ask its `error_format` option for either `'{statistic}'` to get the t-stat, or `'{p.value}'` to get the p-value (or both)
- (you could even ask `export_summs()` for a confidence interval with `error_format = [{conf.low}, {conf.high}]`, by default 95%)
- Both will report "stars" - if the p-value is below a certain $alpha$, you get stars! By default, `*` means $p < .05$, `**` means $p < .01$, and `***` means $p < .001$.

---

# Hypothesis Testing in R

```{r, echo = TRUE}
data(SLID, package = 'carData')
model <- lm(wages ~ education, data = SLID)
summary(model)
```

---

# Hypothesis Testing in R

```{r, echo = TRUE}
export_summs(model, error_format = '(t = {statistic}, p = {p.value})')
```

---

# Hypothesis Testing in R

- We can also see confidence intervals for our coefficients with `


```{r, echo = TRUE}
plot_coefs(model)
```

---

# Hypothesis Testing in R

- How about tests against non-0 nulls, or F tests?
- We can calculate a z-score ourselves and do a non-0 null that way, or we can use `linearHypothesis` from **car** to do F-tests
- Just tell it what value you want to test the coefficient against!

```{r}
library(car)
linearHypothesis(model, c('education = 0'))
linearHypothesis(model, c('education = .75'))
```

Again eyeing the regression table and looking at stars
Using linearHypothesis
(incl. with robust SEs)
Plotting CIs and calculating CIs.


---

# Swirl Practice

Now on to the Hypothesis Testing Swirl lesson!